\thetitle

### 2020-07-08 writing/challenges.tex
revising rational section

The most serious challenge for comparing blame assignment strategies concerns 
the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for empirical research
similar to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a large number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm.  As indicated in the introduction, in
the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of inputs then yields relevant 
data, similar to collecting empirical data in a human-focused discipline.

@@

Another derivative challenge is the problem of mistakes in type
annotations themselves.  Gradual typing enables programmers to move an untyped
code base into the typed realm on an incremental basis. The retroactive addition
of a specific type to a component, even a library, may result from a
misunderstanding of the code and may thus be a mistake itself. Anecdotal 
evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
incorrect-ts, wmwz-ecoop-2017}, and hence must be addressed by any evaluation
method. Conversely, the method not only demands mutators for code that cause
good impedance mismatches, it also calls for a set of mutators that ``break''
type annotations without breaking the code itself.

The second big challenge is the disparity of gradually
typed languages. While the approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate these differences as variables
of a comparison, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems are not
comparable.  For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. The revised transient
variant accommodates many more linguistic features and library calls
in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

Beyond the implementation issue, a comparison also faces the problem that
the three variants---the natural semantics, the transient semantics, and
the erasure semantics---come with three radically different blame
assignment strategies.  For example, the natural semantics assigns blame
to one component while transient assigns blame to a set of components. The
erasure semantics does not blame a component per se, but it comes with an
exception location and the usual stack traces.

Making these strategies comparable is {\em not\/} an issue of language
implementation but the behavior of the rational programmer.
Specifically, the rational programmer must come with \emph{modes}
that represent the alternative interpretations of blame. 
For instance, a programmer using transient may
pick the oldest element of a blame set as the next component to add types because
it corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode does not distinguish between the elements of the
blame set and has the rational programmer type all blamed components
at once. Because this second kind of step 
is more laborious than the first one, a comparison must take 
into account the rational programmer's effort.

Besides the blame information they produce, the three gradual typing
systems also differ in the runtime type checks they perform. Therefore an
application of the evaluation method to a gradual typing system has to
factor in the effect of type checks to determine whether blame helps the
rational programmer find the cause of an impedance mismatch. A solution to this issue
is yet another mode for the rational programmer that uses information only
from the underlying languages exceptions to select the next component(s) to type.
This \emph{exception} mode simulates a rational programmer that ignores
blame and serves as the baseline for blame's value.

>>>>

The most serious challenge for evaluating blame concerns 
the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for empirical research
similar to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a large number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm.  As indicated in the introduction, in
the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of inputs then yields relevant 
data, similar to collecting empirical data in a human-focused discipline.

@@

%Another derivative challenge is the problem of mistakes in type
%annotations themselves.  Gradual typing enables programmers to move an untyped
%code base into the typed realm on an incremental basis. The retroactive addition
%of a specific type to a component, even a library, may result from a
%misunderstanding of the code and may thus be a mistake itself. Anecdotal 
%evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
%incorrect-ts, wmwz-ecoop-2017}, and hence must be addressed by any evaluation
%method. Conversely, the method not only demands mutators for code that cause
%good impedance mismatches, it also calls for a set of mutators that ``break''
%type annotations without breaking the code itself.

The second big challenge is the disparity of gradually
typed languages. While the approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate the effect of these differences 
on its results, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems 
do not share a common baseline and, therefore, do not lend support to any
conclusions about whether the value of blame changes from one system to
another. For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. The revised transient
variant accommodates many more linguistic features and library calls
in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

Beyond the implementation issue, the method also faces the problem that
the three variants---the natural semantics, the transient semantics, and
the erasure semantics---come with three radically different blame
assignment strategies.  For example, the natural semantics assigns blame
to one component while transient assigns blame to a set of components. The
erasure semantics does not blame a component per se, but it comes with an
exception location and the usual stack traces.

Supporting these strategies is {\em not\/} just an issue of language
implementation but affects greatly the behavior of the rational programmer.
Specifically, the rational programmer must come with \emph{modes}
that represent the alternative interpretations of blame. 
For instance, a programmer using transient may
pick the oldest element of a blame set as the next component to add types because
it corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode does not distinguish between the elements of the
blame set and has the rational programmer type all blamed components
at once. Because this second kind of step 
is more laborious than the first one, the analysis of the results must take 
into account the rational programmer's effort.

Besides the blame information they produce, the three gradual typing
systems also differ in the runtime type checks they perform. Therefore an
application of the evaluation method to a gradual typing system has to
factor in the effect of their different type checks to determine whether blame helps the
rational programmer find the cause of an impedance mismatch. A solution to this issue
is yet another mode for the rational programmer that uses information only
from the underlying languages exceptions to select the next component(s) to type.
This \emph{exception} mode simulates a rational programmer that ignores
blame and serves as the baseline for blame's value.

### 2020-07-08 writing/mutate.tex
revising rational section

\section{How to Make Benchmarks} \label{sec:mutate}

@@

>>>>

\section{How to Make Benchmarks for the Value of Blame} \label{sec:mutate}

@@

\begin{figure}
\begin{tabular}{p{1.5cm} | p{10cm} }
  {\bf  name} & {\bf description (author)}  \\

\hline

acquire & Board game simulation (M. Felleisen)  \\%[1em]

\hline
  gregor & Utilities for calendar dates (J. Zeppieri) \\%[1em]

\hline
kcfa & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]

\hline
quadT & Converter from S-expresion source code to PDF format (M. Butterick)\\%[1em]

\hline
quadU & Converter from S-expresion source code to PDF format  (B. Greenman) \\%[1em]

\hline
snake & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]

\hline
suffixtree & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]

\hline
synth & Converter of description of notes and drum beats to WAV format (V. St-Amour \& N. Toronto) \\%[1em]

\hline
take5 & Card game simulator (M.Felleisen)  \\%[1em]

\hline
tetris & Functional implementation of Tetris (D. Van Horn) \\%[1em]

\end{tabular}
  \caption{Benchmarks summary.}
  \label{table:benchmark-descriptions}
\end{figure}

\begin{figure}
  \begin{tabular}{p{2cm} | p{5cm}  | p{5.5cm} }
    {\bf operator} & {\bf description} & {\bf example} \\

\hline
arithmetic 
& Swap an arithmetic operator with another one of the same (or greater)
    arity and a possibly different result type. 
    & \texttt{+} $\rightarrow$ \texttt{-} \\
\hline

boolean 
& Swap \texttt{and} and \texttt{or}. Due to Racket's truthiness it
    possibly changes the type of the operation's result. 
& \texttt{and} $\rightarrow$ \texttt{or} \\

\hline 
constant
& Swap a literal constant with another of the same value but different type.
& \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
\hline 
deletion
    & Deletes the result expression of a \texttt{begin} or \texttt{begin0}
    sequence.
    & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
\hline     
list
    & Swap \texttt{car} and \texttt{cdr}.
    & \texttt{car} $\rightarrow$ \texttt{cdr} \\
 \hline    
class:public
    & Make a public method private and vice versa.
    & 
    \begin{tabular}[t]{@{}l}
      \texttt{(class object\%}\\
        \texttt{\,\,(define/public (m x) x))}\\
       \rightarrow\\ 
      \texttt{(class object\%}\\
        \texttt{\,\,(define/private (m x) x))}\\
    \end{tabular}\\
\hline

class:super
    & Remove \texttt{super-new} calls from class definitions.
    &  \begin{tabular}[t]{@{}l}
      \texttt{(class object\% (super-new))}\\
       \rightarrow\\ 
      \texttt{(class object\% (void))}\\
    \end{tabular}\\
\hline

conditional
    & Negate conditional test expressions.
    &  \begin{tabular}[t]{@{}l}
      \texttt{(cond [(= x 0) 42] ...)}\\ 
       \rightarrow\\ 
      \texttt{(cond [(not (= x 0)) 42] ...)}
    \end{tabular}\\
\hline

class:parent
    & Replace the parent of classes with \texttt{object\%}.
    &  \begin{tabular}[t]{@{}l}
     \texttt{(class foo\% ...)}\\
       \rightarrow\\ 
      \texttt{(class object\% ...)}\\
    \end{tabular}\\
\hline

class:init
  & Swap default values of class fields.
   &  \begin{tabular}[t]{@{}l}
     \texttt{(class object\%}\\
     \texttt{\,\,(field [a 5] [b "hello"]))}\\
       \rightarrow\\ 
      \texttt{(class object\%}\\
     \texttt{\,\,(field [a "hello"] [b 5]))}\\
    \end{tabular}\\
\hline

position
 & Swap the position of subexpressions.
  &  \begin{tabular}[t]{@{}l}
     \texttt{(f a 42 "b" 0)}\\
       \rightarrow\\ 
      \texttt{(f a 42 0 "b")}
    \end{tabular}\\
\hline

class:method
 & Add an extra trivial public method to a class.
 &  \begin{tabular}[t]{@{}l}
   \texttt{(class object\%  ...)}\\
       \rightarrow\\ 
     \texttt{(class object\%}\\
     \texttt{\,\,(define/public (extra x) x)}\\
     \texttt{\,\,...)}
    \end{tabular}\\
\hline

id
 & Swaps a use of a top-level identifier defined with with another defined
 in the same module.
 &\texttt{(f ...)} \rightarrow \texttt{(g ...)}

\end{tabular}
  \caption{Summary of code mutation operators.}
  \label{table:mutation-ops}
\end{figure}

### 2020-07-08 writing/rational.tex
revising rational section

Code migration is the raison d'\^etre of Typed Racket, which is the community
uses ``migratory typing'' instead of gradual typing. Roughly speaking a
programmer uses a gradually typed language to incrementally add types to a code
base. From this perspective, a code base consists of some number of components,
say modules, and a programmer converts a component at a time from its untyped
state to a typed one.

Based on this idea, a metric for migratory paths is a concise way to measure a
rational programmer. The components of the program create a \emph{configuration
space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
of variants of the program, dubbed \emph{configurations}, that differ in their
type specifications. In the setting of migratory typing,\footnote{The method can
be adapted to so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each configuration differs from another in the subset
of typed components. Formally, a program $\system$ is a set of components
$\set{\component}$, and a configuration $\conf$ is a subset of
$\set{\component}$ . These configurations of $\system$ are ordered by the subset
relation and thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$
elements.  The bottom configuration of $\lattice{\system}$ is always $\emptyset$
and the top element is $\system$ itself. In between are all the mixed-type
configurations.

@@

The next three subsections specialize the idea to the various semantics of
gradual typing. Once this machinery is in place, we use the fourth subsection to
state the precise experimental research questions that correspond to the
high-level question on the first page. The final subsection explains how the
rational programmer's effort is measured.

%% -----------------------------------------------------------------------------
\subsection{Modes of Natural Type Programming} \label{sub:natural}

Putting these pieces together:

\begin{quote}
\it 
A natural blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\blame{\system, \conf_i}\}$.
\end{quote}

To define capture whether blame helps rational
programmers in their search for bugs, we define when a trail is successful:
\begin{quote}
\it 
A trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful}  iff its last configuration $\conf_n$ does not type check.
  \end{quote}
  In juxtaposition:
\begin{quote}
\it 
A trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{failing}  iff $\conf_n$ type checks and $\blame{\system,\conf_n}$ is
undefined, meaning $\conf_n$ does not result in blame. 
 \end{quote}
That is failing trails are those that we cannot extend
further following blame and have not reached a configuration that reveals
the bug at
compile time.

 We can now give a precise definition of what it
means for blame to help the rational programmer:
\begin{quote}
\it
  Natural blame is useful for a program $\system$ iff 
  $\lattice{\system}$ contains no failing natural blame trails.
\end{quote}

However, this definition does not give a full picture of how much blame helps the
natural rational programmer. In particular,  a $\lattice{\system}$
may contain no failing natural blame trails because
of the run time type checks natural performs rather than blame, or even as a result
of pure chance. To account for these possibilities, we define two extra
modes for the rational programmer. In the first, the natural programmer
ignores blame and instead decides what component to type next based on
information from the stack trace of the failed runtime type check of a
configuration.  This exception mode of the rational programmer corresponds to \emph{natural
exception trails}:
\begin{quote}
\it 
A natural exception trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$
  where $\error{\system, \conf}$ denotes the component that triggers a
an exception of the underlying language.
\end{quote}

In the second mode, the natural programmer
ignores any kind of information related to the type error and randomly
picks which component to type next:
\begin{quote}
\it 
A natural random trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
that is not in $\conf_i$.
\end{quote}
\noindent

With exception and random trails, we can finaly define the value that blame
adds to the natural semantics: 
\begin{quote}
\it
  Blame adds value to natural for a program $\system$ iff 
  no natural blame trails in  $\lattice{\system}$ are failing while 
 there exist at least one vanilla and one random natural trail that is
  failing.
\end{quote}

\noindent
{\bf Experimental Process.}~Given a set of faulty benchmarks such as the
ones from section~\ref{sec:mutate}, it is straightforward to turn the
above definition into the an experimental question for natural Typed Racket. 
Specifically, for
each benchmark we first construct its configuration lattice. Then, we
determine the configurations of the lattice that detect the fault in the
program. This can happen either because a runtime type check fails or
because Typed Racket's type checker rejects the configuration. In either
case,  each of these configurations is the root of a natural blame trail
in the configuration lattice of the benchmark. We ask whether
\emph{no such natural blame trail is failing}.  Furthermore  we also
examine whether \emph{there are failing exception and random trails that start
from the same roots}.  To answer the questions for a
benchmark, we simply start from our set of roots and extend each
trail according to the corresponding mode of the rational programmer
until no configurations can be added to the trail\footnote{All trails
reach that point since the lattices are finite.}. Finally we check if the
last configuration of each trail type checks or not. If it does the trail
is successful, otherwise it is failing.

%% -----------------------------------------------------------------------------
\subsection{Modes of Optional Type Programming} \label{sub:transient}

The case of erasure semantics is analogous to the third mode of natural.  Since
gradually typed languages with erasure semantics do not come with blame
assignment, a rational programmer can only hope that the underlying safety
checks and their stack traces are helpful.  Thus, in the same veub as for
natural we ask whether the configuration lattice of a given program contains
only successful erasure exception trail and at least one failing erasure random
trail.

%% -----------------------------------------------------------------------------
\subsection{Modes of Transient Type Programming} \label{sub:transient}

Adapting the simulation of rational programmer to transient
requires taking into account the distinguishing
features of each semantics. As we discuss in section~\ref{sec:landscape}, 
transient blames a set of components in contrast to natural's
single-component blame.

For transient, we 
need two transient-specific modes for the rational
programmer instead of natural's blame mode. Specifically, in the first mode, the rational programmer types
all blamed components.  In the second mode, the rational programmer types
the component that is added to the blame set first --- and thus can 
detect an impedance mismatch earlier in the evaluation of a program. 
As before, the two modes give
rise to two different notions of trail:
\begin{quote}
\it 
A  transient all blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i \in \mblame{\system, \conf_i}$
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics; 
\end{quote}
\noindent
and
\begin{quote}
\it 
A transient firstborn  trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
  where $\first{\mblame{\system, \conf}}$ denotes the first component 
  the transient semantics add to the blame set for $\conf$.
\end{quote}

\noindent
{\bf Experimental Process.} Similar to the experiment for natural, we can
test whether blame adds value to transient. However, we must ask two
separate questions about blame; one for each mode of the transient
rational programmer. Specifically, we have two derivative questions: (i)
does blame add value to transient if the rational programmer focuses on all
blamed components of a configuration and (ii) does blame adds value to
transient if the rational programmer focuses on the earliest entry in a
configuration's blame set? The two questions boil down to using 
transient all blame trails and transient
firstborn trails respectively and performing twice the same experiment as that
for natural blame.

%% -----------------------------------------------------------------------------
\subsection{Questions}

In sum, our experiment examines four questions for each benchmark:
\begin{itemize}
\item[$Q_1$] Does natural blame add value to natural?

\item[$Q_2$] Does transient all blame add value to transient?

\item[$Q_3$] Does transient firstborn blame add value to transient?

\item[$Q_*$] Do exceptions add value? For any of the systems? 
\end{itemize}

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different trails/modes of the rational programmer. For example experimental
question $Q_1$ asks whether blame adds value to transient and our experiment
uses the natural blame, exception and random trails to answer it.

@@

@@

In addition to the value of blame, trails permit us to examine the effort a
rational programmer spends to locate a bug.  For instance, the length of a
successful natural blame trail tells us how many components the programmer has
to type before reaching the faulty component. This idea also applies to all
other modes of the rational programmer. The most interesting case is that for
transient all blame trails.  In detail, given a successful transient all blame
trail $\conf_0, ...., \conf_n$ in a configuration lattice $\lattice{\system}$,
the effort of the rational programmer is the number of distinct configurations
in the all blame trails with root $\conf_n$ and length $n$ in
$\lattice{\system}$. Our experiment utilizes this notion of effort to analyze
burden on the rational programmer for finding a bug when the answer to one of
the experimental questions from above is positive for a benchmark.

>>>>

Code migration is the raison d'\^etre of Typed Racket, which is why the community
uses ``migratory typing'' instead of gradual typing. Roughly speaking a
programmer uses a gradually typed language to incrementally add types to a code
base. From this perspective, a code base consists of some number of components,
say modules, and a programmer converts a component at a time from its untyped
state to a typed one.

Based on this idea, a metric for migratory paths is a concise way to
evaluate the effectiveness of a
rational programmer. The components of each program create a \emph{configuration
space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
of variants of the program, dubbed \emph{configurations}, that differ in their
type specifications. In the setting of migratory typing,\footnote{The method can
be adapted to so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each configuration differs from another in the subset
of typed components. Formally, a program $\system$ is a set of components
$\set{\component}$, and a configuration $\conf$ is a subset of
$\set{\component}$ . These configurations of $\system$ are ordered by the subset
relation and thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$
elements.  The bottom configuration of $\lattice{\system}$ is always $\emptyset$
and the top element is $\system$ itself. In between are all the mix-typed
configurations.

@@

The next three subsections specialize the idea to the various semantics of
gradual typing. Once this machinery is in place, we use the fourth subsection to
state the precise experimental research questions that correspond to the
high-level question on the first page and to explain our experimental
process. The final subsection extends the development with a discussion of
how we can measure the rational programmer's effort.

%% -----------------------------------------------------------------------------
\subsection{Modes of the Natural Rational Programmer} \label{sub:natural}

In natural Typed Racket, rational programmers rely on blame when they
attempt to debug a mix-typed programmer; they add type specifications to
the component blame points. In terms of configurations and trails, this
translates to extending a trail with a configuration that adds to the
last configuration of the trail so far the component the latter configuration
blames.  We call trails constructed in this manner \emph{natural blame
trails} and they capture the essence of the natural rational programmer:

\begin{quote}
\it 
A natural blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
  $\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

Natural rational programmers migrate faulty programs along natural blame
trails in order to locate bugs. Specifically, assuming that the types 
they add during migration are correct, when a buggy component migrates
from untyped to typed, the typechecker detects the impedance mismatch
between the broken code of the component and its types and signals a type
error. Hence, trails that end at an ill-typed configuration successfully 
pinpoint the location of bugs:
In formal terms:
\begin{quote}
\it 
A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful}  iff its last configuration $\conf_n$ does not type check.
  \end{quote}
  In juxtaposition:
\begin{quote}
\it 
A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{failing}  iff $\conf_n$ type checks and the trail cannot be extended
  further. 
 \end{quote}
\noindent 
That is failing natural blame trails are those that have not reached a configuration that reveals
the bug at compile time and their last configuration does not produce
blame. Thus the rational programmer receives no hint from the gradual type
system about how to further migrate the faulty program in search for the
bug.

 We can now give a precise definition of what it
means for blame to help the rational programmer:
\begin{quote}
\it
  Natural blame is always useful for a program $\system$ iff 
  $\lattice{\system}$ contains no failing natural blame trails.

   Natural blame is sometimes useful for a program $\system$ iff 
  $\lattice{\system}$ contains successful natural blame trails.

   Natural blame is never useful for a program $\system$ iff 
  $\lattice{\system}$ contains no successful natural blame trails.

\end{quote}

These definitions help a rational programmer gauge whether it pays off to heed blame
when debugging \system. 
However, they do not answer whether blame is a critical piece of 
the rational programmer's process.  For instance, 
typing a random component   could be equally useful as typing the blamed one. 
To account for this situation, we devise a new
mode of the natural rational programmer that 
 follows a random migration process:
\begin{quote}
\it 
A natural random trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
that is not in $\conf_i$.
\end{quote}
\indent
With blame and random natural trails, we can ask whether 
blame adds value over random migration: 
\begin{quote}
\it
  Blame adds value to natural for a program $\system$ over pure chance iff 
  no natural blame trails in  $\lattice{\system}$ are failing while 
 there exists at least one random natural trail that is
  failing.
\end{quote}

Even if the success of the natural rational programmer is not by pure
chance, it could well be because the natural semantics perform runtime type checks at
the right time. In other words, one rational programmer that 
heeds blame could be equally successful as one that only uses information from the exceptions of failed 
runtime type checks. Similar to before, in response we
define the exception mode of the rational programmer:
\begin{quote}
\it 
A natural exception trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$
  where $\error{\system, \conf}$ denotes the component that triggers a
an exception of the underlying language.
\end{quote}
\noindent
We use the new mode to describe what it means for blame to add value over both exceptions and pure
chance:
\begin{quote}
\it
  Blame adds value to natural for a program $\system$ over pure chance and
  exceptions iff no natural blame trails in  $\lattice{\system}$ are failing while 
 there exist at least one random natural trail and one natural exception trail
 that are
  failing.
\end{quote}
\noindent
If blame does add value to natural for a program $\system$ 
over pure chance and exceptions then we can finally
deduce that, for debugging $\system$, blame is a valuable feature of the
natural semantics.

%% -----------------------------------------------------------------------------
\subsection{Modes of the Erasure Rational Programmer} \label{sub:erasure}

 Since gradually typed languages with erasure semantics do not come with blame
assignment, a rational programmer can only hope that the underlying safety
checks and their exceptions are helpful.  Thus, adapting the corresponding
definitions from natural to erasure, we ask whether the effect of erasure
exception in debugging a program $\system$ goes beyond pure chance:

\begin{quote}
\it
  Exceptions adds value to erasure for a program $\system$ over pure chance
  iff no erasure exception trails in  $\lattice{\system}$ are failing while 
 there exists at least one random erasure trail that is failing.
\end{quote}
\noindent

%% -----------------------------------------------------------------------------
\subsection{Modes of the Transient Rational Programmer} \label{sub:transient}

In contrast to erasure, transient does come with blame. As we discuss in
section~\ref{sec:landscape} though, distinctly from natural blame,
transient may blame more than one components. This peculiarity of 
transient raises the question of how the rational programmer should react
when the language produces a blame set. Our answer is that the rational
programmer has least two reasonable options.  The first option is the rational programmer types
all blamed components at once.  In the second option, the rational
programmer selects the component that is added to the blame set first and types
only that --- after all the types of this first component should be able
to  detect an impedance mismatch earlier in the evaluation of a program. 
As before, the two modes give rise to two different notions of trail:
\begin{quote}
\it 
A  transient all blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i = \mblame{\system, \conf_i}$
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics; 
\end{quote}
\noindent
and
\begin{quote}
\it 
A transient firstborn blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
  where $\first{\mblame{\system, \conf}}$ denotes the first component 
  the transient semantics add to the blame set for $\conf$.
\end{quote}
\noindent As for natural we use the transient all blame trails and 
transient firstborn  trails to define when  transient all blame and 
transient firstborn add value to transient over exceptions and pure
chance.

%% -----------------------------------------------------------------------------
\subsection{Experimental Process.}

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket and the 
mutants of section~\ref{sec:mutate}. In particular, our experiment examines 
four questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame is valuable for natural?

\item[$Q_2$] Is all blame valuable for transient?

\item[$Q_3$] Is firstborn blame valuable for transient?

\item[$Q_*$] Are plain exceptions valuable for
  erasure? For the other two systems? 
\end{itemize}

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different trails/modes of the rational programmer. For example experimental
question $Q_1$ asks whether blame is valuable for natural and our experiment
uses the natural blame, exception and random trails to answer it.

@@

In detail, to answer $Q_1$ for a mutant, we first construct the mutant's configuration 
lattice. Then, we determine the configurations of the lattice that detect the fault in the
program. This can happen either because Typed Racket's type checker
outright rejects the configuration or because natural Typed Racket's runtime 
performs a type check that fails. In either
case,  each of these configurations is the root of a trail that we 
extend according the natural blame . 
If no configurations can be added to the trail\footnote{All trails
reach that point since the lattices are finite.}, we check if the
last configuration of the trail type checks or not. If it does the trail
is successful, otherwise it is failing. We repeat the same process for the
roots two more times; once following the natural exception rational programmer and 
once the natural random rational programmer. Finally, we use the
success/failure results for the trails to check if they meet the condition
for natural blame to add value to natural over exceptions and pure chance.
The process is analogues for the other questions and their corresponding
modes of the rational programmer.

@@

In addition to the value of blame, trails permit us to examine the effort a
rational programmer spends to locate a bug. As a metric for the effort, 
we use the number of components the rational programmer has to migrate 
from untyped to typed to find the bug. We can easily and uniformly calculate the value
of the metric for any successful trail in any mode of the rational
programmer (including transient blame all)  as the  size of the difference 
between the trail's root and its final configuration.

Programmer effort can help shed some light to the relative effectiveness of the
three gradual typing systems. For example, consider a configuration of the
lattice of one of our mutants. The configuration corresponds to a
 scenario that the rational programmer attempts to debug. If the
configuration results in blame both under, for instance, the natural and transient semantics,
then the natural blame rational programmer and the two transient blame
rational programmers (all and firstborn) can compete to see which one
reaches a configuration that doesn't type check with less effort. In
general, the effort of trails of different modes that have the same
root is a metric that allows us to compare the effectiveness of these
modes even if they correspond to different gradual typing systems.
Unfortunately, statistically speaking, such comparisons between two modes do not generalize 
beyond the successful trails that share their root in our experiments.
Still they allow us to quantify the difference between the modes of the rational
programmers and thus the different blame assignment strategies  for the specific
debugging scenarios we analyze in our experiment.

### 2020-07-08 writing/sample.tex
revising rational section

>>>>

As we mention above we treat each definitions as a component. A
consequence of this decision is that examining the whole lattice
for each of the mutants becomes unfeasible.
As figure~\ref{table:benchmarks} shows,  most of the programs contain dozens of definitions  
resulting in lattices with as many as $1.41 \times 10^{28}$ configurations.
Since running a single configuration can take 10 minutes or more 
(due to contract checking and instrumentation), running every possible configuration is impractical.
Evidence from similar contexts suggests that randomly sampling the lattice instead of
an exhaustive exploration is an effective alternative~\cite{greenman2019evaluate}.
Hence we randomly sample enough significant configurations
from the lattice to obtain a confidence of 0.95 (with margin of error
0.05) about whether our hypotheses hold or not.
See accompanying appendix for the details of the calculations for the
sufficient size of samples for this estimate.
Exploring more of the lattice would yield higher confidence in the generalizability
of our results for each benchmark in cases where no violations of a property
(e.g. Blame Trail) are found.
In such cases, we could be more certain that we did not simply miss
the configuration(s) that reveal the violation by exploring the lattice more fully.
Our choice of random sample counts reflects the (informal) standard practice of estimating
results to 95 percent or higher confidence.
